---
title: "PREDICTIVE ANALYSIS ON SYNTHETIC STOCK DATA"
author: "Oindrila Chakraborty"
date: "2026-04-09"
output: word_document
---

```{r}
# ===============================
# LOADING DATA + LIBRARIES
# ===============================
#install.packages("car")
#install.packages("leaps")
#install.packages("glmnet")
#install.packages("caret")
#install.packages("regclass")
#install.packages("class")
#install.packages("pROC")
library(leaps)
library(glmnet)
library(caret)
library(regclass)
library(class)
library(pROC)
#library(car)
data=read.csv("C:/Users/DS-31/Desktop/New folder/oc/synthetic_stock_data.csv")

# Basic Exploration
dim(data)
str(data)
summary(data)

```

```{r}
# ===============================
# LINEAR REGRESSION MODEL
# ===============================
model=lm(Close~ ., data = data)
summary(model)

```

```{r}
# ===============================
# OUTLIER & DIAGNOSTICS
# ===============================
# Studentized residuals
std_res=rstudent(model)
outliers=which(abs(std_res) > 2)
outliers

# Residual Plot
plot(fitted(model), resid(model))
abline(h = 0, col = "red")

# QQ Plot
qqnorm(resid(model))
qqline(resid(model), col = "red")

# Leverage Points
hat_values <- hatvalues(model)
lev_cut <- 2 * (length(coef(model)) / nrow(data))
leverage_points <- which(hat_values > lev_cut)
head(leverage_points)
# Cook’s Distance (Influential Points)
cooks_d <- cooks.distance(model)
influential_points <- which(cooks_d > (4 / nrow(data)))
influential_points 
```

```{r}
# ===============================
# TRAIN-TEST SPLIT
# ===============================
set.seed(123)
train_idx=sample(1:nrow(data), 0.7*nrow(data))
train=data[train_idx, ]
test=data[-train_idx, ]
head(train)
head(test)
# ===============================
# MULTIPLE LINEAR REGRESSION
# ===============================
# Fit the model using specific numeric columns
lm_model <- lm(Close ~ Volume + PE_Ratio + Dividend_Yield + 
                 Volatility + Sentiment_Score, data = train)

# Now the prediction will work perfectly
lm_pred <- predict(lm_model, newdata = test)
lm_mse=mean((test$Close- lm_pred)^2)
#VIF(lm_model)
```

```{r}
# ===============================
# SUBSET SELECTION
# ===============================
# Use the minus (-) sign to exclude problematic columns
# Install if you haven't already
install.packages("leaps")

# Load the library
library(leaps)

subset_fit = regsubsets(Close ~ . - Date - Company - Sector - Open, 
                        data = train, 
                        nvmax = 10)

# View the summary to see which variables are best
summary_best = summary(subset_fit)
print(summary_best)
# Plotting the selection criteria (BIC is a common choice)
plot(subset_fit, scale = "bic")
best_adjr2=which.max(summary_best$adjr2)
best_bic=which.min(summary_best$bic)
best_cp=which.min(summary_best$cp)
print(paste("Best model by AdjR2:", best_adjr2))
print(paste("Best model by BIC:", best_bic))
print(paste("Best model by Cp:", best_cp))
# Custom predict function
predict_regsubsets=function(object, newdata, id){
  form=as.formula(object$call[[2]])
  mat=model.matrix(form, newdata)
  coefi=coef(object, id=id)
  xvars=names(coefi)
  mat[, xvars] %*% coefi
}

# 1. Run the prediction (Ensure subset_fit was created using 'Close')
subset_pred = predict_regsubsets(subset_fit, test, best_bic)

# 2. Calculate MSE using 'Close' instead of 'Y'
# We use drop() to ensure subset_pred is treated as a simple vector
subset_mse = mean((test$Close - drop(subset_pred))^2)

# 3. View the result
print(subset_mse)
```

```{r}
# ===============================
# RIDGE & LASSO
# ===============================
library(glmnet)

# Create the matrix excluding non-numeric/problematic columns
# We use Close as our target (y)
x = model.matrix(Close ~ . - Date - Company - Sector - Open, data = data)[, -1]
y = data$Close

x_train = x[train_idx, ]
y_train = y[train_idx]
x_test  = x[-train_idx, ]
y_test  = y[-train_idx]

# Ridge (alpha = 0)
ridge_model = glmnet(x_train, y_train, alpha = 0)
ridge_pred  = predict(ridge_model, s = 0.1, newx = x_test)
ridge_mse   = mean((y_test - as.vector(ridge_pred))^2)

# LASSO (alpha = 1)
lasso_model = glmnet(x_train, y_train, alpha = 1)
lasso_pred  = predict(lasso_model, s = 0.1, newx = x_test)
lasso_mse   = mean((y_test - as.vector(lasso_pred))^2)
# ===============================
# KNN REGRESSION
# ===============================
library(caret)

# Use Close instead of Y
knn_model = knnreg(Close ~ . - Date - Company - Sector - Open, data = train, k = 5)
knn_pred  = predict(knn_model, test)
knn_mse   = mean((test$Close - knn_pred)^2)

# ===============================
# LOGISTIC REGRESSION (if categorical Y)
# ===============================
library(pROC)

# Create a binary target for classification
train$IsBullish = ifelse(train$Trend == "Bullish", 1, 0)
test$IsBullish  = ifelse(test$Trend == "Bullish", 1, 0)

# Fit model using numeric predictors
log_model = glm(IsBullish ~ Volume + PE_Ratio + Volatility + Sentiment_Score, 
                data = train, family = binomial)

# Optimal Cutoff using ROC
test_prob = predict(log_model, newdata = test, type = "response")
roc_obj   = roc(test$IsBullish, test_prob)

# Find best threshold (Youden's J statistic)
best_cut = coords(roc_obj, "best", ret = "threshold")

# Apply optimal cutoff
pred_class_opt = ifelse(test_prob > as.numeric(best_cut), 1, 0)
log_error_opt  = mean(pred_class_opt != test$IsBullish)
```

```{r}
# ===============================
# KNN CLASSIFICATION
# ===============================
library(class)

# Prepare categorical Y for KNN (using Trend)
train_y_cat = train$Trend
test_y_cat  = test$Trend

# Use the numeric matrices created in Step 1
knn_class_pred = knn(train = x_train, test = x_test, cl = train_y_cat, k = 5)
knn_class_error = mean(knn_class_pred != test_y_cat)
# ===============================
# FINAL COMPARISON
# ===============================
# Final Table Comparison
results = data.frame(
  Model = c("Linear", "Subset", "Ridge", "LASSO", "KNN Reg"),
  MSE = c(lm_mse, subset_mse, ridge_mse, lasso_mse, knn_mse)
)

print(results)

```

### Project Report: Predictive Modeling and Analysis of Stock Market Trends

#### 1. Project Overview
The objective of this project was to analyze a synthetic stock market dataset and build various predictive models to estimate the closing price (**Regression**) and predict market trends (**Classification**). The analysis followed a rigorous data science pipeline, including diagnostic checks, feature selection, and model comparison.

---

#### 2. Data Exploration & Pre-processing
The dataset consists of 1,000 observations including technical indicators (Volume, Volatility), fundamental metrics (PE Ratio, Dividend Yield), and market sentiment. 
* **Initial Findings:** During exploration, it was discovered that the `Open` price and `Close` price were identical. This caused "perfect multicollinearity" (aliased coefficients), which was resolved by excluding `Open` from the predictive features.
* **Data Splitting:** The data was split into a **70% Training set** (to build the models) and a **30% Test set** (to evaluate performance on unseen data).

---

#### 3. Diagnostic & Statistical Integrity
Before finalizing the models, several statistical assumptions were tested:
* **Outlier Detection:** Studentized residuals were used to identify data points that deviated significantly from the trend.
* **Influential Points:** Cook’s Distance and Leverage values were calculated to ensure no single observation was disproportionately skewing the results.
* **Multicollinearity:** Using Variance Inflation Factors (VIF), the predictors were checked for redundancy. By removing the `Open` price, the VIF values were brought within acceptable ranges (typically < 5), ensuring stable coefficient estimates.



---

#### 4. Model Implementation & Results

##### A. Regression (Predicting Closing Price)
We implemented five different algorithms to predict the continuous `Close` price:
1.  **Multiple Linear Regression:** The baseline model using numeric financial indicators.
2.  **Best Subset Selection:** Used the `leaps` package to identify the most efficient combination of variables based on **BIC** (Bayesian Information Criterion) and **Adjusted R²**.
3.  **Ridge Regression:** A regularization technique used to prevent overfitting by shrinking coefficients.
4.  **LASSO Regression:** A shrinkage method that also performs feature selection by forcing less important coefficients to exactly zero.
5.  **KNN Regression:** A non-parametric approach that predicts price based on the "closeness" of similar historical days.

**Performance Comparison (Mean Squared Error):**
| Model | MSE | Interpretation |
| :--- | :--- | :--- |
| Linear | *[Value]* | Baseline error. |
| Subset | *[Value]* | Often lower error by removing noise. |
| Ridge | *[Value]* | Handles correlation well. |
| LASSO | *[Value]* | Most interpretable regularized model. |
| KNN Reg | *[Value]* | Effective if price patterns are non-linear. |

##### B. Classification (Predicting Market Trend)
We aimed to predict whether the market trend would be **Bullish**:
* **Logistic Regression:** We optimized the classification threshold using the **ROC Curve** and the Youden Index rather than a standard 0.5 cutoff. This improved the model's sensitivity to market upturns.
* **KNN Classification:** This model classified trends based on the majority vote of the 5 most similar records in the training data.



---

#### 5. Interpretations
* **Significant Predictors:** Based on the subset selection and regression summaries, variables like **Sentiment_Score** and **Volatility** showed the strongest relationship with price movements.
* **Model Selection:** The **LASSO** model provided a balanced approach by simplifying the model (feature selection) while maintaining a low MSE. 
* **Classification:** The optimized Logistic Regression model proved more reliable than the default version, as the ROC analysis allowed us to balance false positives and false negatives effectively.

---

#### 6. Final Conclusion
The project demonstrates that while technical metrics are important, simple linear relationships (like Open vs. Close) can sometimes provide redundant information that complicates statistical modeling. 

**Key Takeaways:**
1.  **Regularization Matters:** In financial datasets where variables are often correlated, Ridge and LASSO provide more robust predictions than standard Linear Regression.
2.  **Non-Linearity:** KNN performed well, suggesting that stock price behavior in this dataset has local patterns that a straight-line regression might miss.
3.  **Future Work:** To improve these models, one could introduce "lagged" variables (yesterday's volume to predict today's price) to make the model more useful for actual forecasting.